Joint Neighborhood Subgraphs Link Prediction

### 2017-08-03 paper.tex
Add files via upload

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Dinh Tran-Van$^1$ \and Alessandro Sperduti$^1$ \and Fabrizio Costa$^2$}
%
\authorrunning{Dinh Tran Van, Alessandro Sperduti, and Fabrizio Costa$^2$}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract}
A crucial computational task for relational and network data is the ''link prediction problem'' which allows for example to discover unknown interactions between proteins to explain the mechanism of a disease in biological networks, or to suggest novel products for a customer in a e-commerce recommendation system. Most approaches however do not effectively exploit the contextual information available in the neighborhood of each edge. Here we propose to cast the problem as a binary classification task over the union of the pair of subgraphs located at the endpoints of each edge. We model the classification task using a support vector machine endowed with an efficient graph kernel and achieve state-of-the-art results on several benchmark datasets.

% Graphs are common data structure used to represent data in which the relation between entities are encoded. A number of graph-based methods and systems have been proposed, varying from biology to recommendation to social network. Most methods have to take partially observed graphs as the input to proceed due to the lack of information. This prevents them from showing promising results. Link prediction becomes an effective paradigm to solve this problem. However, most existing link prediction methods are based on the transitive manner which cannot effectively exploit graph structures. Here we propose a method that first represents each link between two nodes by a graph composed of their neighborhood subgraphs. We then employ a decompositional graph kernel to measure the similarity between links. An empirical experiment on different datasets proves that our proposed method shows the state of the art for link prediction.
\keywords{Link prediction, graph kernels}
\end{abstract}

\section{Introduction and related work}
We are witnessing a constant increase of the rate at which data is being produced and made available in machine readable formats. Interestingly it is not only the quantity of data that is increasing, but also its complexity, i.e. not only are we measuring a number of attributes or features for each data point, but we are also capturing their mutual relationships, that is, we are considering non independent and identically distributed (non i.i.d.) data. This yields collections that are best represented as graphs or relational data bases and requires a more complex form of analysis. As cursory examples of application domains that are particularly  it is sufficient to mention social networks, where nodes are people and edges encode a type of association such as friendship or co-authorship, bioinformatics, where nodes are proteins and metabolites and edges represent a type of chemical interaction such as catalysis or signaling, and e-commerce, where nodes are people and goods and edges encode a ''buy'' or ''like'' relationship.
A key characteristic of this type of data collections is the sparseness and dynamic nature, i.e. the fact that the number of recorded relations is significantly smaller than the number of all possible pairwise relations, and the fact that these relations evolve in time. A crucial computational task is then the ''link prediction problem'' which allows to suggest friends, or possible collaborators for scientists in social networks, or to discover unknown interactions between proteins to explain the mechanism of a disease in biological networks, or to suggest novel products to be bought to a customer in a e-commerce recommendation system. The many approaches to link prediction that exist in literature can be partitioned according to 1) whether additional or ''side'' information is available for nodes and edges or rather only the network topology is considered and 2) whether the approach is unsupervised or supervised.

Unsupervised methods are non-adaptive (i.e. they do not have parameters that are tuned on the specific problem instance), and can therefore be computationally efficient. In general they define a score for any node pair that is proportional to the existence likelihood of an edge between the two nodes. \textit{Adamic-Adar} \cite{adamic} computes the weighted sum over the common neighbors where the weight is inversely proportional to the (log of) each neighbor node degree. The \textit{preferential attachment} method computes a score simply as the product of the node degrees in an attempt to exploit the 'rich get richer' property of certain network dynamics. \textit{Katz} \cite{katz} takes into account the number of common paths with different lengths between two nodes, assigning more weight to shorter paths. The \textit{Leicht-Holme-Newman} method \cite{lhni} computes the number of intermediate nodes. In \cite{matrix-factorization} the score is derived from the singular value decomposition of the adjacency matrix.

@@

In general supervised methods exhibit better accuracies compared to unsupervised methods although incurring in much higher computational and memory complexity costs.
Moreover, most approaches implicitly represent the link prediction problem and the inference used to tackle it as a disjunction over the edges, that is, information on edges is propagated in such a fashion so that for a node to have $k$ neighbors or $k+1$ does not make a drastic difference.
We claim that this hypothesis is likely putting a cap on the discriminative power of classifiers and therefore we propose a novel supervised method that employs a conjunctive representation. We call the method ''joint neighborhood subgraphs link prediction'' (JNSL). The key idea here is to transform the link prediction task into a binary classification on suitable small subgraphs which we then solve using an efficient graph kernel method.

@@

\subsection{Definitions and notation}
We represent a problem instance as a graph $G=(V,E)$ where $V$ is the set of nodes and $E$ is the set of links. The set $E$ is partitioned into the subset of observed links ($O$) and the subset of unobserved links ($U$). Like other approaches we assume that all unobserved links are indeed ''non-links'' and we therefore define the link prediction problem as the task of ranking candidate links from the most to the least probable to recover links in $O$ but not in $U$ exploiting only the network topology.

We define the \textit{distance} $\mathcal{D}(u,v)$ between two nodes $u$ and $v$, as the number of edges on the shortest path between them. The \textit{neighborhood} of a node $u$ with radius $r$, $N_r(u) = \lbrace v\ |\ d(u,v) \leq r \rbrace$, is the set of nodes at distance no greater than $r$ from $u$. The corresponding \textit{neighborhood subgraph} $\mathcal{N}_{r}^{u}$ is the  subgraph induced by the neighborhood (i.e. considering all the edges with endpoints in $N_r(u)$). The \textit{degree} of a node $u$, $d(u) = |\mathcal{N}_{1}^{u}|$, is the cardinality of its neighborhood. The maximum node degree in the graph $G$ is $d(G)$.

\subsection{Link encoding as subgraphs union}
\label{sec:link}
Most methods for link prediction compute pairwise nodes similarities treating  the nodes defining the candidate edge independently. Instead we propose to jointly consider both candidate endpoint nodes together with their extended ''context''. To do so we build a graph starting from the two nodes and the underlying network. Given nodes $u$ and $v$, we first extract the two neighborhood sets with a user defined radius $R$ rooted at $u$ and $v$ to obtain $N_R(u)$ and $N_R(v)$, respectively. We then consider the graph $\mathcal{J}$ induced by the set union $N_R(u) \cup N_R(u)$. Finally we add an auxiliary node $w$ and the necessary edges to connect it to $u$ and $v$ (see Fig.\ref{fig:example}).

@@

\subsection{Node labeling}
We propose to use a graph kernel approach to classify the subgraphs encoding each link. In our setup nodes are not endowed with any ''side'' information. However to increase the discriminative power of the similarity notion induced by the graph kernel, instead of assuming a dummy, non-informative label on each node, we propose to use a node labeling function $\ell$ which assigns as the discrete label the node degree. More precisely, for nodes having degree less than or equal than a user defined threshold $T$ ($T=5$ in our experimental evaluation) we use as label the degree value. Degree values larger than $T$ are subsequently discretized into $k$ levels. Here the implicit assumption is that nodes with similar degrees have common properties. Formally, the labeling function is defined as:
\begin{center}
$\ell(u) = \left\{
	\begin{array}{ll}
		d(u),\  & \mbox{if } d(u) \leq T \\
		T+i,\ & \mbox{if } d(u) > T
	\end{array}
\right.$,
\end{center}
where $i = \lceil \frac{d(u)-T}{bin}\rceil$, $bin = \frac{d(G)-T}{\lambda - T}$ and $\lambda$ ($\lambda > T$) is the maximum number of symbols used. The value of $\lambda$ depend on the degree distribution and can be tuned as a hyperparameter of the approach.

\subsection{The graph kernel}
Here we briefly describe an efficient graph kernel called the Neighborhood Subgraph Pairs Distance kernel (NSPDK) introduced in \cite{nspdk}. NSPDK is an instance of ''decompositional'' kernels \cite{convolution-kernel} based on the idea of counting the number of common small subgraphs between two graphs. The subgraphs are pairs of neighborhoods whose roots are at a short distance.

@@

We evaluate the performance of employed methods splitting 10 times the data in a train and a test part. For \textit{Protein}, \textit{Metabolic}, \textit{Nips} and \textit{Conflict} networks, we use 10$\%$ of the edges to induce the training set while for \textit{Condmat} and \textit{Powergrid} we use 90$\%$ of the links. The performance of each method is computed as the average of the AUC-ROC over the 10 rounds.

@@

\section{Results and Discussion}
In table \ref{result_table} we report the performance of link prediction methods measured as the AUC-ROC value on 6 datasets. From the results on the table, we can group methods into two groups based on their performances: supervised methods and unsupervised methods. The performance of supervised methods are considerably higher than unsuperivsed ones in most cases, except in the Conflict dataset where Sup-Top outperforms Fact+Scores, but with a very small difference. Concerning supervised methods, JNSL outperforms Fact-Scores in all cases. The difference between their performance is small in PowerGrid and Protein datasets with 0.5$\%$ and 0.8$\%$, respectively. And the big gap is in the Condmat dataset with 7.4$\%$.

\begin{table}
\caption{AUC-ROC performance on 6 datasets. Legend: AA: Adamic-Adar, PA \cite{pa} preferential Attachment, SHP: Shortest Path, Sup-Top: Liear regression running on unsupervised scores \cite{matrix-factorization}:, SVD \cite{matrix-factorization}:: Singular value decomposition, Fact+Scores \cite{matrix-factorization}: Factorization with unsupervised scores, JNSL: joint neighborhood subgraph link (our method). In bold the highest score.}
\centering
\setlength{\tabcolsep}{1mm}
\begin{tabular}{|c|c|c|c|c|c|c|}

@@

\section{Conclusion and Future Work}
We have presented a novel approach to link prediction in absence of side information that can effectively exploit the topological contextual information available in the neighborhood of each edge. We have empirically shown that this approach achieves very competitive results compared to other state-of-the-art methods.
In future work, we will investigate how to make use of multiple and heterogeneous information sources when these are available for nodes and edges.

>>>>

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Dinh Tran-Van$^1$ \and Alessandro Sperduti$^1$ \and Fabrizio Costa$^2$}
%
\authorrunning{Dinh Tran-Van, Alessandro Sperduti, and Fabrizio Costa$^2$}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract}
A crucial computational task for relational and network data is the ``link prediction problem'' which allows for example to discover unknown interactions between proteins to explain the mechanism of a disease in biological networks, or to suggest novel products for a customer in a e-commerce recommendation system. Most link prediction approaches however do not effectively exploit the contextual information available in the neighborhood of each edge. Here we propose to cast the problem as a binary classification task over the union of the pair of subgraphs located at the endpoints of each edge. We model the classification task using a support vector machine endowed with an efficient graph kernel and achieve state-of-the-art results on several benchmark datasets.

\keywords{Link prediction, graph kernels}
\end{abstract}

\section{Introduction and related work}
We are witnessing a constant increase of the rate at which data is being produced and made available in machine readable formats. Interestingly it is not only the quantity of data that is increasing, but also its complexity, i.e. not only are we measuring a number of attributes or features for each data point, but we are also capturing their mutual relationships, that is, we are considering non independent and identically distributed (non i.i.d.) data. This yields collections that are best represented as graphs or relational data bases and requires a more complex form of analysis. As cursory examples of application domains that are social networks, where nodes are people and edges encode a type of association such as friendship or co-authorship, bioinformatics, where nodes are proteins and metabolites and edges represent a type of chemical interaction such as catalysis or signaling, and e-commerce, where nodes are people and goods and edges encode a ``buy'' or ``like'' relationship.
A key characteristic of this type of data collections is the sparseness and dynamic nature, i.e. the fact that the number of recorded relations is significantly smaller than the number of all possible pairwise relations, and the fact that these relations evolve in time. A crucial computational task is then the ``link prediction problem'' which allows to suggest friends, or possible collaborators for scientists in social networks, or to discover unknown interactions between proteins to explain the mechanism of a disease in biological networks, or to suggest novel products to be bought to a customer in a e-commerce recommendation system. Many approaches to link prediction that exist in literature can be partitioned according to \textit{i}) whether additional or ``side'' information is available for nodes and edges or rather only the network topology is considered and \textit{ii}) whether the approach is unsupervised or supervised.

Unsupervised methods are non-adaptive (i.e. they do not have parameters that are tuned on the specific problem instance), and can therefore be computationally efficient. In general they define a score for any node pair that is proportional to the existence likelihood of an edge between the two nodes. \textit{Adamic-Adar} \cite{adamic} computes the weighted sum over the common neighbors where the weight is inversely proportional to the (log of) each neighbor node degree. The \textit{preferential attachment} method computes a score simply as the product of the node degrees in an attempt to exploit the ``rich get richer'' property of certain network dynamics. \textit{Katz} \cite{katz} takes into account the number of common paths with different lengths between two nodes, assigning more weight to shorter paths. The \textit{Leicht-Holme-Newman} method \cite{lhni} computes the number of intermediate nodes. In \cite{matrix-factorization} the score is derived from the singular value decomposition of the adjacency matrix.

@@

In general supervised methods exhibit better accuracies compared to unsupervised methods although incurring in much higher computational and memory complexity costs.
Moreover, most approaches implicitly represent the link prediction problem and the inference used to tackle it as a disjunction over the edges, that is, information on edges is propagated in such a fashion so that for a node to have $k$ neighbors or $k+1$ does not make a drastic difference.
We claim that this hypothesis is likely putting a cap on the discriminative power of classifiers and therefore we propose a novel supervised method that employs a conjunctive representation. We call the method ``joint neighborhood subgraphs link prediction'' (JNSL). The key idea here is to transform the link prediction task into a binary classification on suitable small subgraphs which we then solve using an efficient graph kernel method.

@@

\subsection{Definitions and notation}
We represent a problem instance as a graph $G=(V,E)$ where $V$ is the set of nodes and $E$ is the set of links. The set $E$ is partitioned into the subset of observed links ($O$) and the subset of unobserved links ($U$). Like other approaches we assume that all un-observed links are indeed ``non-links'' and we therefore define the link prediction problem as the task of ranking candidate links from the most to the least probable to recover links in $O$ but not in $U$ exploiting only the network topology.

We define the \textit{distance} $\mathcal{D}(u,v)$ between two nodes $u$ and $v$, as the number of edges on the shortest path between them. The \textit{neighborhood} of a node $u$ with radius $r$, $N_r(u) = \lbrace v\ |\ \mathcal{D}(u,v) \leq r \rbrace$, is the set of nodes at distance no greater than $r$ from $u$. The corresponding \textit{neighborhood subgraph} $\mathcal{N}_{r}^{u}$ is the  subgraph induced by the neighborhood (i.e. considering all the edges with endpoints in $N_r(u)$). The \textit{degree} of a node $u$, $d(u) = |\mathcal{N}_{1}^{u}|$, is the cardinality of its neighborhood. The maximum node degree in the graph $G$ is $d(G)$.

\subsection{Link encoding as subgraphs union}
\label{sec:link}
Most methods for link prediction compute pairwise nodes similarities treating  the nodes defining the candidate edge independently. Instead we propose to jointly consider both candidate endpoint nodes together with their extended ``context''. To do so we build a graph starting from the two nodes and the underlying network. Given nodes $u$ and $v$, we first extract the two neighborhood sets with a user defined radius $R$ rooted at $u$ and $v$ to obtain $N_R(u)$ and $N_R(v)$, respectively. We then consider the graph $\mathcal{J}$ induced by the set union $N_R(u) \cup N_R(u)$. Finally we add an auxiliary node $w$ and the necessary edges to connect it to $u$ and $v$ (see Fig.\ref{fig:example}).

@@

\subsection{Node labeling}
We propose to use a graph kernel approach to classify the subgraphs encoding each link. In our setup nodes are not endowed with any ``side'' information. However to increase the discriminative power of the similarity notion induced by the graph kernel, instead of assuming a dummy, non-informative label on each node, we propose to use a node labeling function $\ell$ which assigns as the discrete label the node degree. More precisely, for nodes having degree less than or equal than a user defined threshold $T$ ($T=5$ in our experimental evaluation) we use as label the degree value. Degree values larger than $T$ are subsequently discretized into $k$ levels. Here the implicit assumption is that nodes with similar degrees have common properties. Formally, the labeling function is defined as:
\begin{center}
$\ell(u) = \left\{
	\begin{array}{ll}
		d(u),\  & \mbox{if } d(u) \leq T \\
		T+i,\ & \mbox{if } d(u) > T
	\end{array}
\right.$,
\end{center}
where $i = \lceil \frac{d(u)-T}{bin}\rceil$, $bin = \frac{d(G)-T}{\lambda - T}$ and $\lambda$ ($\lambda > T$) is the maximum number of symbols used. The value of $\lambda$ depends on the degree distribution and can be tuned as a hyperparameter of the approach.

\subsection{The graph kernel}
Here we briefly describe an efficient graph kernel called the Neighborhood Subgraph Pairs Distance kernel (NSPDK) introduced in \cite{nspdk}. NSPDK is an instance of ``decompositional'' kernels \cite{convolution-kernel} based on the idea of counting the number of common small subgraphs between two graphs. The subgraphs are pairs of neighborhoods whose roots are at a short distance.

@@

We evaluate the performance of employed methods by splitting 10 times the data in a train and a test part. For \textit{Protein}, \textit{Metabolic}, \textit{Nips} and \textit{Conflict} networks, we use 10$\%$ of the edges to induce the training set while for \textit{Condmat} and \textit{Powergrid} we use 90$\%$ of the links. The performance of each method is computed as the average of the AUC-ROC over the 10 rounds.

@@

\section{Results and discussion}
In Table \ref{result_table}, we report the performance of link prediction methods measured as the AUC-ROC value on 6 datasets. From the results on the table, we can group methods into two groups based on their performances: supervised methods and unsupervised methods. The performance of supervised methods are considerably higher than unsuperivsed ones in most cases, except in the Conflict dataset where Sup-Top outperforms Fact+Scores, but with a very small difference. Concerning supervised methods, JNSL outperforms Fact-Scores in all cases. The difference between their performance is small in PowerGrid and Protein datasets with 0.5$\%$ and 0.8$\%$, respectively. And the big gap is in the Condmat dataset with 7.4$\%$.

\begin{table}
\caption{AUC-ROC performance on 6 datasets. Legend: AA: Adamic-Adar, PA \cite{pa} preferential Attachment, SHP: Shortest Path, Sup-Top \cite{matrix-factorization}: Liear regression running on unsupervised scores, SVD \cite{matrix-factorization}: Singular value decomposition, Fact+Scores \cite{matrix-factorization}: Factorization with unsupervised scores, JNSL: joint neighborhood subgraph link (our method). In bold the highest score.}
\centering
\setlength{\tabcolsep}{1mm}
\begin{tabular}{|c|c|c|c|c|c|c|}

@@

\section{Conclusion and future work}
We have presented a novel approach to link prediction in absence of side information that can effectively exploit the topological contextual information available in the neighborhood of each edge. We have empirically shown that this approach achieves very competitive results compared to other state-of-the-art methods.
In future work, we will investigate how to make use of multiple and heterogeneous information sources when these are available for nodes and edges.
